# Old content about randomness (2025)

## Estimating the mean

...

### Across variance of the data

Having seen how the variance of our estimates depend on the sample size, let's also quickly see how they depend on the distribution of the data.

We will write a new simulator function, which has first argument `var`, the variance of the data:

```{r}
simulate_estimating_means_2 <- \(var, nreps, n, mean) {
  replicate(nreps, {
    mean(rnorm(n=n, mean=mean, sd=sqrt(var)))
  })
}
```

Run this simulation for five values of variance:

```{r}
n = 1000
var_values <- 1:5
est_means <- lapply(
  var_values,
  simulate_estimating_means_2,
  nreps=nreps,
  n=n,
  mean=true_mean
)
```

As expected, the variance of the estimate increases with the variance of the data:

```{r box_means_by_var}
names(est_means) <- paste0("var=", var_values)
boxplot(
  est_means,
  range = 0,
  main = "estimated means over data variance",
  xlab = "data variance"
)
```

The variance of the estimate is simply linear with the variance of the data:

```{r plot_var_means_by_var}
var_of_means <- sapply(est_means, var)
plot(
  var_values,
  var_of_means,
  type = "b",
  main = "estimate variance over data variance",
  xlab = "data variance"
)
# theoretical variance:
points(var_values, var_values / n, type = "b", col = "blue")
```

### Detection power

Statistical power for testing is essentially asking if our estimate is away from a particular value (say 0, or some other experimentally-motivated value) relative to the variance of our estimate. 

We can demonstrate this with a simple example, asking if our estimate of the mean is far from a threshold value.

Let's set up our simulator with different sample sizes, from 2 to 100.

```{r}
# is mean greater than a number?
n_values <- c(2,5,1:10*10)
est_means <- lapply(
  n_values,
  simulate_estimating_means,
  nreps=nreps,
  mean=true_mean,
  sd=true_sd
)
```

We compute power (also called "sensitivity"), by asking how often our estimate will be far from the threshold value. It makes sense to define "far" by a value derived from the _estimate variance_. That way, we are seeing if we are farther away from a value that would be expected given that the estimate we are computing has a given spread.

We will tally below how often the mean is 2 times the standard deviation of our estimate away from the threshold. 

::: {.callout-note collapse="false"}
## Question

Are we computing or estimating power?
:::


```{r plot_detected_by_n}
est_sd <- sqrt(1/n_values) # theoretical estimate SD
thr <- 4.5 # some threshold
detected <- sapply(seq_along(est_means), {
  \(i) mean(est_means[[i]] > 2 * est_sd[i] + thr)
  })
plot(n_values, detected, type="b", log="x", 
     main="power over sample size")
abline(h = 0.5, lwd=2, lty=2, col="grey")
```

::: {.callout-note collapse="false"}
## Question

Above we used the _theoretical variance_ of the estimate, which we could compute because we knew that the true variance of the data was 1. In reality what would we need to do in such a setting?
:::

In many experiments, we don't specify what kind of an effect size would be interesting. We often just compare some effect to 0 (a null hypothesis of "no effect"). This often works in practice, especially with experiments with low to moderate sample size (say n=3 replicates per group), because the estimate variance is much larger than the different between 0 and a truly biologically relevant value.

However, with large sample size, and high replicate correlation, we may need to motivate a non-zero effect size of interest, or risk many or even the majority of times detecting an "interesting" difference from 0.

This has been discussed in the RNA-seq literature, including the DESeq2 paper.

## Estimating variance

We conclude this note by repeating the above simulator, but this time using it to estimate the _variance_ of data. We change the last line of our simulator function to compute `var`:

```{r}
simulate_estimating_vars <- \(n, nreps, mean, sd) {
  replicate(nreps, {
    var( rnorm(n=n, mean=mean, sd=sd) )
  })
}
```

And we iterate over a number of sample sizes:

```{r}
true_sd <- 2
n_values <- c(3, 5, 10, 50, 100)
est_vars <- lapply(
  n_values,
  simulate_estimating_vars,
  nreps=nreps,
  mean=true_mean,
  sd=true_sd
)
```

Note how large the estimated variances can be when n=3. We don't have very stable estimates of the variance until much higher than n=10.

These are the (squared) values that we would need to use in the detection context above, as we wouldn't have the true variance of the data to compute the standard deviation of the estimate of the mean.

```{r box_vars_by_n}
names(est_vars) <- paste0("n=", n_values)
boxplot(
  est_vars,
  range = 0,
  main = "estimated variance over sample size",
  xlab = "simulation sample size"
)
```

As before, we can explore the mean and variance of the estimates, now of the _variance_ estimates.

```{r plot_vars_by_n}
(mean_of_vars <- sapply(est_vars, mean))
(var_of_vars <- sapply(est_vars, var))
plot(n_values, var_of_vars, type="b", log="xy")
# theoretical variance:
points(n_values, 2 * true_sd^4 / (n_values - 1), type="b", col="blue")
```
