---
title: "Tidy Data"
author: "Michael Love"
date: "July 2025"
license: "CC-BY"
format:
  html:
    toc: true
    code-fold: false
    code-tools: true
    highlight-style: github
    code-line-numbers: false
    embed-resources: true
params:
  skip_answers: true
---

## Introduction

*Tidy data* is a standard way of organizing data values within a dataset (Wickham 2014). A dataset is tidy when:

1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

Most real-world data arrives in an untidy form. The *tidyr* package provides tools to reshape data into tidy format, after which the *dplyr* package provides a consistent grammar for transformation and summarization.

In this notebook we will work with toy datasets to practice identifying tidy vs. untidy structure, and use `pivot_longer()`, `pivot_wider()`, and core *dplyr* verbs to tidy and wrangle data.

## What does untidy data look like?

A common form of untidy data is a "wide" table where values of one variable are spread across column names. Consider a table of gene expression counts where each column is a sample:

```{r}
library(tidyr)
library(dplyr)

counts_wide <- tibble(
  gene    = c("GAPDH", "IL6", "TNF"),
  naive   = c(450, 120, 280),
  IFNg    = c(430, 230, 210),
  SL1344  = c(460, 180, 340)
)
counts_wide
```

Here `naive`, `IFNg`, and `SL1344` are values of a `condition` variable, not variables themselves. To make this tidy we need `pivot_longer()`.

## `pivot_longer()`

`pivot_longer()` takes multiple columns and collapses them into two: one for the former column names and one for the values.

```{r}
counts_long <- counts_wide |>
  pivot_longer(
    cols      = -gene,       # every column except 'gene'
    names_to  = "condition",
    values_to = "count"
  )
counts_long
```

::: {.callout-note collapse="false"}
## Question

How many rows does `counts_long` have compared to `counts_wide`? Why?
:::

## `pivot_wider()`

The inverse operation is `pivot_wider()`, which spreads a key-value pair back into separate columns. This is sometimes useful for display or for feeding data into a matrix-based function.

```{r}
counts_long |>
  pivot_wider(
    names_from  = condition,
    values_from = count
  )
```

## Exercise: pivot a table

::: {.callout-note collapse="false"}
## Exercise

The `penguins` dataset (from *palmerpenguins*) has four body measurement columns: `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g`. Pivot those four columns to long format with columns `species`, `measurement`, and `value`, keeping only `species` as the identifier.
:::

```{r}
#| exercise: pivot_ex
#| setup: true
library(tidyr)
library(dplyr)
library(palmerpenguins)
```

```{r}
#| caption: Pivot penguins measurements to long format
#| exercise: pivot_ex
penguins |>
  select(species, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  pivot_longer(
    cols      = ______,
    names_to  = ______,
    values_to = ______
  )
```

::: {.solution exercise="pivot_ex"}

#### Solution

`-species` selects every column except `species`, which is the identifier.

```{r}
#| exercise: pivot_ex
#| solution: true
penguins |>
  select(species, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  pivot_longer(
    cols      = -species,
    names_to  = "measurement",
    values_to = "value"
  )
```

:::

```{r}
#| exercise: pivot_ex
#| check: true
gradethis::grade_this_code()
```

## Core dplyr verbs

With tidy data in hand, *dplyr* provides five key verbs:

| Verb | Purpose |
|---|---|
| `filter()` | Keep rows matching a condition |
| `select()` | Keep (or drop) columns |
| `mutate()` | Add or modify columns |
| `summarize()` | Reduce rows to a summary |
| `arrange()` | Sort rows |

These are usually combined with `group_by()` to apply operations within groups.

```{r}
counts_long |>
  filter(condition != "naive") |>
  group_by(condition) |>
  summarize(total = sum(count), mean_count = mean(count))
```

## Exercise: filter and summarize

::: {.callout-note collapse="false"}
## Exercise

Using `counts_long`, keep only rows where `count > 100`, then compute the mean count per `condition`.
:::

```{r}
#| exercise: filter_ex
#| setup: true
library(tidyr)
library(dplyr)

counts_long <- counts_wide |>
  pivot_longer(cols = -gene, names_to = "condition", values_to = "count")
```

```{r}
#| caption: Filter and summarize counts_long
#| exercise: filter_ex
counts_long |>
  filter(______) |>
  group_by(______) |>
  summarize(mean_count = ______)
```

::: {.solution exercise="filter_ex"}

#### Solution

```{r}
#| exercise: filter_ex
#| solution: true
counts_long |>
  filter(count > 100) |>
  group_by(condition) |>
  summarize(mean_count = mean(count))
```

:::

```{r}
#| exercise: filter_ex
#| check: true
gradethis::grade_this_code()
```

## Plotting tidy data with base R

One payoff of tidy data is that a single column can drive an aesthetic like color. With `counts_long` already in long format, we can map the `condition` column directly to point color.

```{r}
gene_levels <- c("GAPDH", "IL6", "TNF")
conditions  <- c("naive", "IFNg", "SL1344")
colors      <- c("dodgerblue", "firebrick", "forestgreen")
names(colors) <- conditions

plot(
  as.numeric(factor(counts_long$gene, levels = gene_levels)),
  counts_long$count,
  col  = colors[counts_long$condition],
  pch  = 16,
  xaxt = "n",
  xlab = "gene",
  ylab = "count",
  main = "counts by gene and condition"
)
axis(1, at = 1:3, labels = gene_levels)
legend(
  "topright", legend = conditions, col = colors, 
  pch = 16,  y.intersp = 1.5, inset = 0.05
  )
```

The key line is `col = colors[counts_long$condition]`: because `condition` is a character vector of condition names, and `colors` is a named vector, this indexes `colors` by name to produce a color for every row.


## Session info

```{r}
sessionInfo()
```
