---
title: "Tidy Data"
author: "Michael Love"
date: "July 2025"
license: "CC-BY"
format:
  html:
    toc: true
    code-fold: false
    code-tools: true
    highlight-style: github
    code-line-numbers: false
    embed-resources: true
params:
  skip_answers: true
---

## Introduction

*Tidy data* is a standard way of organizing data values within a dataset (Wickham 2014). A dataset is tidy when:

1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

Most real-world data arrives in an untidy form. The *tidyr* package provides tools to reshape data into tidy format, after which the *dplyr* package provides a consistent grammar for transformation and summarization.

In this notebook we will work with toy datasets to practice identifying tidy vs. untidy structure, and use `pivot_longer()`, `pivot_wider()`, and core *dplyr* verbs to tidy and wrangle data.

## What does untidy data look like?

A common form of untidy data is a "wide" table where values of one variable are spread across column names. Consider a table of gene expression counts where each column is a sample. The three genes are:

- **GAPDH** — glycolytic enzyme expressed in nearly all cells; widely used as a housekeeping reference gene because its expression is constitutive and stable.
- **IL6** — pro-inflammatory cytokine that activates acute-phase responses; strongly induced by bacterial infection through NF-κB signaling.
- **IRF1** — transcription factor and direct target of the IFNg–JAK–STAT1 axis; a canonical marker of the interferon response.

```{r}
library(tidyr)
library(dplyr)

counts_wide <- tibble(
  gene    = c("GAPDH", "IL6", "IRF1"),
  control   = c(520, 30, 50),
  IFNg    = c(560, 90, 290),
  Salmonella  = c(500, 320, 310)
)
counts_wide
```

Here `control`, `IFNg`, and `Salmonella` are values of a `condition` variable, not variables themselves. To make this tidy we need `pivot_longer()`.

## `pivot_longer()`

`pivot_longer()` takes multiple columns and collapses them into two: one for the former column names and one for the values.

```{r}
counts_long <- counts_wide |>
  pivot_longer(
    cols      = -gene,       # every column except 'gene'
    names_to  = "condition",
    values_to = "count"
  )
counts_long
```

::: {.callout-note collapse="false"}
## Question

How many rows does `counts_long` have compared to `counts_wide`? Why?
:::

## `pivot_wider()`

The inverse operation is `pivot_wider()`, which spreads a key-value pair back into separate columns. This is sometimes useful for display or for feeding data into a matrix-based function.

```{r}
counts_long |>
  pivot_wider(
    names_from  = condition,
    values_from = count
  )
```

## Exercise: pivot a table

::: {.callout-note collapse="false"}
## Exercise

The `penguins` dataset (from *palmerpenguins*) has four body measurement columns: `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g`. Pivot those four columns to long format with columns `species`, `measurement`, and `value`, keeping only `species` as the identifier.
:::

```{r}
library(palmerpenguins)
penguins
```

```{r}
#| caption: Pivot penguins measurements to long format
#| exercise: pivot_ex
#| eval: false
penguins |>
  select(species, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  pivot_longer(
    cols      = ______,
    names_to  = ______,
    values_to = ______
  )
```

::: {.solution exercise="pivot_ex"}

#### Solution

`-species` selects every column except `species`, which is the identifier.

```{r}
#| exercise: pivot_ex
#| solution: true
penguins |>
  select(species, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  pivot_longer(
    cols      = -species,
    names_to  = "measurement",
    values_to = "value"
  )
```

:::

```{r}
#| exercise: pivot_ex
#| check: true
#| eval: false
gradethis::grade_this_code()
```

## Core dplyr verbs

With tidy data in hand, *dplyr* provides five key verbs:

| Verb | Purpose |
|---|---|
| `filter()` | Keep rows matching a condition |
| `select()` | Keep (or drop) columns |
| `mutate()` | Add or modify columns |
| `summarize()` | Reduce rows to a summary |
| `arrange()` | Sort rows |

These are usually combined with `group_by()` to apply operations within groups.

```{r}
counts_long |>
  filter(condition != "control") |>
  group_by(condition) |>
  summarize(total = sum(count), mean_count = mean(count))
```

## Exercise: filter and summarize

::: {.callout-note collapse="false"}
## Exercise

Using `counts_long`, keep only rows where `count > 100`, then compute the mean count per `condition`.
:::

```{r}
#| caption: Filter and summarize counts_long
#| exercise: filter_ex
#| eval: false
counts_long |>
  filter(______) |>
  group_by(______) |>
  summarize(mean_count = ______)
```

::: {.solution exercise="filter_ex"}

#### Solution

```{r}
#| exercise: filter_ex
#| solution: true
counts_long |>
  filter(count > 100) |>
  group_by(condition) |>
  summarize(mean_count = mean(count))
```

:::

```{r}
#| exercise: filter_ex
#| check: true
#| eval: false
gradethis::grade_this_code()
```

## Plotting tidy data with `plot()`

One payoff of tidy data is that a single column can drive an aesthetic like color. With `counts_long` already in long format, we can map the `condition` column directly to point color.

```{r}
gene_levels <- counts_wide$gene
conditions <- names(counts_wide)[-1]
colors <- palette.colors(7)[c(7,4,3)]
names(colors) <- conditions

plot(
  as.numeric(factor(counts_long$gene, levels = gene_levels)),
  counts_long$count,
  col  = colors[counts_long$condition],
  pch  = 16,
  cex = 1.5,
  xaxt = "n",
  xlab = "gene",
  ylab = "count",
  main = "counts by gene and condition"
)
axis(1, at = 1:3, labels = gene_levels)
legend(
  "topright", legend = conditions, col = colors, 
  pch = 16, inset = 0.02
  )
```

The key line is `col = colors[counts_long$condition]`: because `condition` is a character vector of condition names, and `colors` is a named vector, this indexes `colors` by name to produce a color for every row.


## Plotting tidy data with ggplot2

The same plot requires far less code with *ggplot2*, and many things that take manual effort in base R — like the legend — appear automatically as sensible defaults.

```{r}
library(ggplot2)

ggplot(
  data = counts_long, 
  mapping = aes(x = gene, y = count, color = condition)
) +
  geom_point(size = 3)
```

The `color = condition` mapping is all it takes: ggplot2 assigns colors and adds a *guide* (legend) automatically. In base R we had to build a named color vector and call `legend()` by hand.

*ggplot2* is built around a layered grammar of graphics. Each `+` adds a new layer or modification — a geometry (`geom_point`), a scale, a theme, a facet — so you can start with a minimal plot and layer complexity on top without rewriting anything. For instance, to facet by condition we can simply add `facet_wrap(~ condition)`:

```{r}
ggplot(
  data = counts_long,
  mapping = aes(x = gene, y = count, color = condition)
) +
  geom_point(size = 3) +
  facet_wrap(~ condition)
```

Later we will show how to add *statistical summary layers* — such as group means with `stat_summary()` or regression slopes with `geom_smooth()` — by dropping in one more line.

## Tips for making data findable, accessible, interoperable, and re-usable

Above we considered the choice of tidy data vs other formats of storing or presenting data. Here we describe some other tips that help make data more easy to use. 

A well-known set of guiding principles of working with and sharing data is called 
[FAIR](https://www.go-fair.org/fair-principles/), an acronym representing the words in the section header here.

Some tips for making data both tidy and FAIR are as follows:

  - Missing values are data. While typically represented with `NA` in R or `NaN` in python, one can model missingness itself, converting to a binary "shadow" matrix. See [Tidy Missing Data](https://naniar.njtierney.com/articles/naniar.html#tidy-missing-data-the-shadow-matrix) in the _naniar_ package vignette. Avoid `-1` for missing or flagged as this can lead to downstream confusion.
  - Dates, units, and categories should be explicit and consistent (no "10mg" mixed with "10"). Units can be provided in the column name (`dosage_mg`). It's a good idea to provide an accompanying data dictionary defining units and categories. Use standard formats for dates such as `ISO 8601: YYYY-MM-DD`. For working in R, see the [lubridate](https://lubridate.tidyverse.org/) package.
  - One thing per cell. Avoid embedding multiple values (e.g., "120/80" for blood pressure). Discourage use of highlighting, color, or style in place of data; instead use an additional column taking boolean values, e.g. `flagged` or `highlight`.
  - Column names should always be machine-readable: no spaces, no special characters, unambiguous. Replace dashes with underscore. When not sure what to name a column, try to find a previously published dataset and re-use their naming scheme rather than invent a new one.
  - Raw data is sacred: never overwrite it. Keep a clean separation between raw and processed data. Raw data can be excluded from version control, but should be described including provenance in an accompanying `README.md`. Where did this data come from, how was it collected? Ideally provide a script for downloading from stable, remote sources such as Zenodo.
  - Finally, a stranger, or a future "you", should be able to re-run everything. Comments, links, software versions (see "Session info" below), can be of great value for future data and analysis re-use. If using randomness as part of analysis (e.g. bootstrapping, sub-sampling, etc.) always set a random seed. For simulations, keep track of the seeds used to generate the multiple simulated datasets.


## Preserving raw data without overwriting

A common operation in genomics is to center (and potentially scale) each gene's expression values by subtracting the gene's mean across conditions (perhaps after taking the logarithm). One tempting approach works directly on a matrix and reassigns the result:

```{r}
counts_mat <- as.matrix(counts_wide[,-1])
rownames(counts_mat) <- counts_wide[,1,drop=TRUE]
counts_mat <- log10(counts_mat + 1)
10^(rowMeans(counts_mat)) # geometric mean
counts_mat <- counts_mat - rowMeans(counts_mat)
counts_mat
```

This works numerically, but the reassignment destroys information. The values in `counts_mat` are no longer counts — they are deviations of original counts — yet the variable name gives no indication of what has been done. More importantly, the original information about the scale of the counts per gene is gone: we can no longer recover that GAPDH had ~530 counts, while IL6 had ~100 and IRF1 had ~160, in terms of the geometric mean.

The tidy alternative is to leave `counts_long` untouched and add new columns with `mutate()`, keeping the original `count` alongside intermediate values. A reader (or a future you) can immediately see both the absolute level and the deviation, and the operation is self-documenting in the column names.

::: {.callout-note collapse="false"}
## Exercise

Starting from `counts_long`, add two new columns without overwriting anything:

- `log_count`: log10 of the count plus one
- `ctr_log10_count`: `log_count` centered by subtracting the per-gene mean
:::

```{r}
#| caption: Add log_count and ctr_log10_count to counts_long
#| exercise: mutate_ex
#| eval: false
counts_long |>
  ______(log_count = ______) |>
  ______(______) |>
  ______(ctr_log10_count = ______)
```

::: {.solution exercise="mutate_ex"}

#### Solution

```{r}
#| exercise: mutate_ex
#| solution: true
counts_long |>
  mutate(log_count = log10(count + 1)) |>
  group_by(gene) |>
  mutate(ctr_log10_count = log_count - mean(log_count))
```

:::

```{r}
#| exercise: mutate_ex
#| check: true
#| eval: false
gradethis::grade_this_code()
```

## Exercise: modeling missingness

The FAIR tips above note that missing values are data and that missingness itself can be modeled. Here we construct a dataset where missingness is *not at random* — larger penguins are more likely to have their body mass unrecorded — then explore whether we can detect that pattern from the data alone.

We start from the complete cases of the `penguins` dataset, then simulate a realistic missing-not-at-random (MNAR) mechanism: penguins heavier than 5500 g have their body mass value suppressed 75% of the time. A companion indicator column records whether each value is missing.

```{r}
library(palmerpenguins)

set.seed(42)
penguins_complete <- penguins |> drop_na()

penguins_obs <- penguins_complete |>
  mutate(
    body_mass_missing = body_mass_g > 5500 & runif(n()) < 0.75,
    body_mass_obs     = if_else(body_mass_missing, NA_real_, body_mass_g)
  )

penguins_obs |>
  select(species, body_mass_g, body_mass_obs, body_mass_missing) |>
  filter(body_mass_g > 5000)
```

## Session info

`sessionInfo()` prints the R version, operating system, locale, and the name and version of every package loaded in the current session. Including it at the end of a script or notebook is a lightweight form of provenance: if a result changes or a bug appears months later, the session info is often the first place to look — it tells you exactly what software produced the output. It is also essential for reproducibility reports and as supplementary information in publications.

A popular alternative is `sessionInfo()` from the *sessioninfo* package, which formats packages in a cleaner table and records the *source* of each package — CRAN, Bioconductor, or a GitHub repository with its commit SHA — which matters when packages installed from GitHub are not on a numbered release. *devtools* depends on *sessioninfo* and re-exports the same function as `devtools::session_info()`, so if devtools is already loaded you can call it either way.

```{r}
sessionInfo()
```

```{r}
devtools::session_info()
```
